---
title: "Supplementary Materials Tables"
author: "Benjamin Shear"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

## Packages

```{r}
library(tidyverse)
library(vtable)
library(kableExtra)
```

## Functions

```{r}
# add summary stats at bottom of CNT-SUB-YEAR tables
add_summary_rows <- function(x) {
  
  stats <- bind_cols(
    CNT=c("Mean","SD","Min","Max"),
    bind_rows(
      apply(x[,-1],2,mean),
      apply(x[,-1],2,sd),
      apply(x[,-1],2,min),
      apply(x[,-1],2,max)
    ))

  bind_rows(x,stats)
  
}
```

# Load Data

```{r}
d <- readRDS(file = "2d-cog-scored-long-allyears-standard-subset.rds") %>%
  arrange(CNT, YEAR, SUB, STUID, item_id)

d <- mutate(d,
            FEMALE=ifelse(GENDER=="FEMALE",1,0))

cnt_names <- read.csv(file = "0-CNT-codes.csv")
d <- left_join(d, cnt_names, by="CNT")
```

# Table S3. Item Properties Summary

```{r}
item_data <- d %>%
  mutate(is_mc=ifelse(is_mc==1,"MC","CR")) %>%
  group_by(SUB, YEAR, item_id, is_mc, subscale, process_short, text_format) %>%
  summarise() %>%
  ungroup()

items_all <- item_data %>%
  group_by(SUB,YEAR,is_mc) %>%
  summarise(N=n()) %>%
  mutate(property="All") %>%
  pivot_wider(names_from = c("YEAR","is_mc"), values_from="N")

math_subscale <- item_data %>%
  filter(SUB=="M") %>%
  rename(property=subscale) %>%
  group_by(SUB,YEAR,is_mc,property) %>%
  summarise(N=n()) %>%
  pivot_wider(names_from = c("YEAR","is_mc"), values_from="N")

math_process <- item_data %>%
  filter(SUB=="M") %>%
  rename(property=process_short) %>%
  group_by(SUB,YEAR,is_mc,property) %>%
  summarise(N=n()) %>%
  pivot_wider(names_from = c("YEAR","is_mc"), values_from="N")

read_process <- item_data %>%
  filter(SUB=="R") %>%
  rename(property=process_short) %>%
  group_by(SUB,YEAR,is_mc,property) %>%
  summarise(N=n()) %>%
  pivot_wider(names_from = c("YEAR","is_mc"), values_from="N")

read_textformat <- item_data %>%
  filter(SUB=="R") %>%
  rename(property=text_format) %>%
  group_by(SUB,YEAR,is_mc,property) %>%
  summarise(N=n()) %>%
  pivot_wider(names_from = c("YEAR","is_mc"), values_from="N")
```

```{r}
bind_rows(
  filter(items_all, SUB=="M"),
  math_subscale,
  math_process,
  filter(items_all, SUB=="R"),
  read_process,
  read_textformat
) %>%
  kable() %>% kable_styling()
```

# Table S5. List of countries and languages

```{r}
d %>%
  group_by(CNT, Country, YEAR, TESTLANG) %>%
  summarise() %>%
  pivot_wider(names_from="YEAR", values_from="TESTLANG") %>%
  kable() %>% kable_styling()
```

# Table S6 Contents

## Sample sizes by CNT-YEAR-SUB

```{r}
d %>%
  group_by(CNT, YEAR, SUB) %>%
  summarise(N=length(unique(STUID))) %>%
  arrange(CNT, SUB, YEAR) %>%
  pivot_wider(names_from = c("SUB","YEAR"), values_from="N")  %>%
  arrange(CNT) %>%
  add_summary_rows() %>%
  kable() %>% kable_styling()
```

## Number of items by CNT-SUB-YEAR

```{r}
d %>%
  group_by(CNT, YEAR, SUB) %>%
  summarise(N=length(unique(item_id))) %>%
  arrange(CNT, SUB, YEAR) %>%
  pivot_wider(names_from = c("SUB","YEAR"), values_from="N")  %>%
  arrange(CNT) %>%
  ungroup() %>%
  add_summary_rows() %>%
  kable(digits=2) %>% kable_styling()
```

## Average item p-values by CNT-SUB-YEAR

```{r}
item_pvals <- d %>%
  group_by(CNT, YEAR, SUB, item_id) %>%
  summarise(P=mean(ypart)) %>%
  group_by(CNT, YEAR, SUB) %>%
  summarise(P=mean(P)) %>%
  arrange(CNT, SUB, YEAR)

item_pvals %>%
  pivot_wider(names_from = c("SUB","YEAR"), values_from="P")  %>%
  arrange(CNT) %>%
  add_summary_rows() %>%
  kable(digits=2) %>% kable_styling()
```

## Cohen's d based on observed scores by CNT-SUB-YEAR

```{r}
cohen_d_vals <- d %>%
  group_by(CNT, SUB, YEAR, STUID, GENDER) %>%
  summarise(X=mean(ypart)) %>%
  group_by(CNT, SUB, YEAR) %>%
  summarise(
    d = (mean(X[GENDER=="FEMALE"])-mean(X[GENDER=="MALE"]))/
      ((sd(X[GENDER=="FEMALE"])+sd(X[GENDER=="MALE"]))/2)
  )

cohen_d_vals %>%
  pivot_wider(names_from = c("SUB","YEAR"), values_from="d") %>%
  add_summary_rows() %>%
  kable(digits=2) %>% kable_styling()
```

## Item format effects observed scores by CNT-SUB-YEAR

```{r}
dformat_vals <- d %>%
  group_by(CNT, SUB, YEAR, STUID, GENDER, is_mc) %>%
  summarise(X=mean(ypart)) %>%
  group_by(CNT, SUB, YEAR, is_mc) %>%
  summarise(
    d = (mean(X[GENDER=="FEMALE"])-mean(X[GENDER=="MALE"]))/
      ((sd(X[GENDER=="FEMALE"])+sd(X[GENDER=="MALE"]))/2)
  ) %>%
  pivot_wider(names_from = c("is_mc"), values_from="d") %>%
  mutate(deff = `0`-`1`) %>%
  select(-`0`, -`1`) %>%
  ungroup()
  
dformat_vals %>%
  pivot_wider(names_from = c("SUB","YEAR"), values_from="deff") %>%
  add_summary_rows() %>%
  kable(digits=2) %>% kable_styling()
```

# Table S7. Combined summary table all CNT

```{r}
item_cols <- d %>%
  group_by(SUB,YEAR,item_id,is_mc) %>%
  summarise(
    x=mean(ypart)
  ) %>%
  group_by(SUB,YEAR) %>%
  summarise(
    NAll=n(),
    NMC=sum(is_mc),
    NCR=sum(is_mc==0),
    PAll=mean(x),
    PMC=mean(x[is_mc==1]),
    PCR=mean(x[is_mc==0])
  ) %>% ungroup()

student_cols <- d %>%
  group_by(SUB,YEAR,CNT,STUID,FEMALE,AGE,GRADE) %>%
  summarise() %>%
  group_by(SUB,YEAR) %>%
  summarise(
    N=n(),
    `%Female`=mean(FEMALE),
    MnAge=mean(AGE),
    MnGrd=mean(GRADE)
  )

# Next, summarize differences in percent correct scores, limiting to students who responded to at least 1 MC and 1 CR

perc_cor_diffs <- d %>%
  group_by(CNT,SUB,YEAR,STUID,FEMALE) %>%
  summarise(
    N=n(),
    NMC=sum(is_mc),
    NCR=sum(is_mc==0),
    All=mean(ypart),
    MC=mean(ypart[is_mc==1]),
    CR=mean(ypart[is_mc==0])
  ) %>%
  filter(N>=5,NMC>=1,NCR>=1) %>%
  pivot_longer(c("All","MC","CR")) %>%
  group_by(SUB,YEAR,name) %>%
  summarise(
    d=(mean(value[FEMALE==1])-mean(value[FEMALE==0]))/
      ((sd(value[FEMALE==1])+sd(value[FEMALE==0]))/2),
    N=n(),
  ) %>%
  pivot_wider(names_from="name",values_from=c("d","N")) %>%
  mutate(d_Diff=d_CR-d_MC)

# Now combine columns together. 

left_join(item_cols, student_cols) %>%
  select(SUB,YEAR,NAll,NCR,NMC,PAll,PCR,PMC,N,`%Female`) %>%
  pivot_longer(cols = -c("SUB","YEAR")) %>%
  unite(subyear, SUB,YEAR) %>%
  pivot_wider(names_from = "subyear", values_from="value") %>%
  bind_rows(., perc_cor_diffs %>%
              select("SUB","YEAR", starts_with("d_")) %>%
              pivot_longer(cols = -c("SUB","YEAR")) %>%
              unite(subyear, SUB,YEAR) %>%
              pivot_wider(names_from = "subyear", values_from="value")) %>%
  arrange(match(name,c(
    "NAll","NCR","NMC",
    "PAll","PCR","PMC",
    "d_All","d_CR","d_MC", "d_Diff",
    "N","%Female"))) %>%
  kable(digits=3) %>% kable_styling()
```

# Extra figures

```{r}
combined_vals <- left_join(item_pvals, cohen_d_vals) %>%
  left_join(., dformat_vals)
```

```{r}
combined_vals %>%
  ggplot(aes(x=d, y=deff)) +
  geom_point() +
  facet_grid(SUB~YEAR)
```

```{r}
combined_vals %>%
  ggplot(aes(x=d, y=P)) +
  geom_point() +
  facet_grid(SUB~YEAR)
```

```{r, fig.height=10, fig.width=10}
combined_vals %>%
  pivot_longer(cols = c("P","d","deff")) %>%
  ggplot(aes(x=value)) +
  geom_histogram(fill="grey", color="black") +
  facet_grid(SUB+YEAR~name, scales="free_x")
```

```{r, fig.height=5, fig.width=10}
combined_vals %>%
  pivot_longer(cols = c("P","d","deff")) %>%
  ggplot(aes(x=CNT, y=value)) +
  geom_point() +
  facet_grid(name~SUB+YEAR)
```


# Sensitivity checks.

## Item Format Effects - full credit scoring

1=full credit; 0=no or part credit

```{r}
format_effects_full <- d %>%
  group_by(CNT, SUB, YEAR, STUID, GENDER, is_mc) %>%
  summarise(X=mean(yfull)) %>%
  group_by(CNT, SUB, YEAR, is_mc) %>%
  summarise(
    d = (mean(X[GENDER=="FEMALE"])-mean(X[GENDER=="MALE"]))/
      ((sd(X[GENDER=="FEMALE"])+sd(X[GENDER=="MALE"]))/2)
  ) %>%
  pivot_wider(names_from = c("is_mc"), values_from="d") %>%
  mutate(eff = `0`-`1`) %>%
  select(-`0`, -`1`)

format_effects_full %>%
  pivot_wider(names_from = c("SUB","YEAR"), values_from="eff") %>%
  add_summary_rows() %>%
  kable(digits=2) %>% kable_styling()
```

## Item Format Effects - partial credit scoring

1=part or full credit; 0=no credit

```{r}
format_effects_part <- d %>%
  group_by(CNT, SUB, YEAR, STUID, GENDER, is_mc) %>%
  summarise(X=mean(ypart)) %>%
  group_by(CNT, SUB, YEAR, is_mc) %>%
  summarise(
    d = (mean(X[GENDER=="FEMALE"])-mean(X[GENDER=="MALE"]))/
      ((sd(X[GENDER=="FEMALE"])+sd(X[GENDER=="MALE"]))/2)
  ) %>%
  pivot_wider(names_from = c("is_mc"), values_from="d") %>%
  mutate(eff = `0`-`1`) %>%
  select(-`0`, -`1`)

format_effects_part %>%
  pivot_wider(names_from = c("SUB","YEAR"), values_from="eff") %>%
  add_summary_rows() %>%
  kable(digits=2) %>% kable_styling()
```

## Item Format Effects - random groups

Robustness check - use a random grouping variable instead of GENDER. How do item format effects look then?

```{r}
set.seed(112822)

d$RUNIF <- round(runif(n = nrow(d)))

format_effects_rand <- d %>%
  mutate(
    GENDER_R = ifelse(RUNIF==1,"FEMALE","MALE")
  ) %>%
  group_by(CNT, SUB, YEAR, STUID, GENDER_R, is_mc) %>%
  summarise(X=mean(ypart)) %>%
  group_by(CNT, SUB, YEAR, is_mc) %>%
  summarise(
    d = (mean(X[GENDER_R=="FEMALE"])-mean(X[GENDER_R=="MALE"]))/((sd(X[GENDER_R=="FEMALE"])+sd(X[GENDER_R=="MALE"]))/2)
  ) %>%
  pivot_wider(names_from = c("is_mc"), values_from="d") %>%
  mutate(eff = `0`-`1`) %>%
  select(-`0`, -`1`)

format_effects_rand %>% 
  pivot_wider(names_from = c("SUB","YEAR"), values_from="eff") %>%
  add_summary_rows() %>%
  kable(digits=2) %>% kable_styling()
```

## Summary Comparison

```{r}
format_effects_full$type <- "fullcredit"
format_effects_part$type <- "partcredit"
format_effects_rand$type <- "randomgrp"
bind_rows(format_effects_full,
          format_effects_part,
          format_effects_rand) %>%
  group_by(type, SUB) %>%
  summarise(
    N=n(),
    Mn=mean(eff),
    sd=sd(eff)
  ) %>%
  pivot_wider(names_from="SUB", values_from=c("Mn","sd"))
```


