---
title: "Summary and regression models for US Rasch DIF Estimates"
author: "Benjamin Shear"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    code_folding: hide
---

# Setup

```{r}
library(tidyverse)
library(kableExtra)
library(TAM)
library(vtable)
library(sjPlot)
library(metafor)
```

# Load data

```{r}
dif <- read.csv(file = "4a-usa-tam-dif-ests-ypart.csv")
```

# Prepare data for regression analysis

```{r}
dif <- dif %>%
  mutate(
    
    is_cr = 1-is_mc,
    year12=ifelse(year==2012,1,0),
    year15=ifelse(year==2015,1,0),
    
    # create dummies for summary stats
    sub_quant=ifelse(subscale=="Quantity",1,0),
    sub_data=ifelse(subscale=="Uncertainty and data",1,0),
    sub_space=ifelse(subscale=="Space and Shape",1,0),
    sub_chng=ifelse(subscale=="Change and Relationships",1,0),
    
    proc_emp=ifelse(process_short=="Employ",1,0),
    proc_form=ifelse(process_short=="Formulate",1,0),
    proc_int_m=ifelse(process_short=="Interpret",1,0),
    
    proc_int_r=ifelse(process_short=="Integrate and interpret",1,0),
    proc_ref=ifelse(process_short=="Reflect and evaluate",1,0),
    proc_acc=ifelse(process_short=="Access and retrieve",1,0),
    
    txt_cont=ifelse(text_format=="Continuous",1,0),
    txt_mix=ifelse(text_format%in%c("Mixed","Multiple"),1,0),
    txt_non=ifelse(text_format=="Non-continuous",1,0)
  ) %>%
  arrange(subject, year, item_id) %>%
  group_by(subject) %>%
  mutate(
    xsi_c=xsi-mean(xsi),
  ) %>% ungroup() %>%
  select(
    subject, year, item_id,
    year12, year15,
    theta_sd, main_eff,
    int_pval, is_cr, is_mc,
    xsi_dif, xsi_dif_se,
    xsi, xsi_se, xsi_c,
    dif, dif_se,
    subscale, process_short, text_format,
    starts_with("sub_"),
    starts_with("proc_"),
    starts_with("txt_")
  )
```

```{r}
saveRDS(dif, file = "4b-dif-estimates-data-US-ypart.RDS")
```

Add in SE of the left out item with DIF equal to sum of all other items.

```{r}
dif <- dif %>%
  group_by(subject, year) %>%
  mutate(
    xsi_dif_se = ifelse(xsi_dif_se==0, sqrt(sum(xsi_dif_se^2)), xsi_dif_se),
    dif_se=ifelse(dif_se==0, xsi_dif_se*2, dif_se)
  ) %>% ungroup()
```

# Boxplot of DIF estimates

```{r}
dif %>%
  mutate(
    format=ifelse(is_mc==1,"MC","CR"),
    sub=ifelse(subject=="M","Math","Reading"),
    DIF=dif
  ) %>%
  ggplot(aes(x=format, y=DIF)) +
  geom_hline(yintercept = 0) +
  geom_boxplot(color="grey", outlier.shape = NA, width=0.3) +
  geom_jitter(color="black", size=1, alpha=0.6, width=0.02) +
  facet_grid(sub~year) +
  xlab("") + 
  theme_bw(base_size = 14)
```

# Boxplot of DIF, with significance

```{r}
dif %>%
  mutate(
    format=ifelse(is_mc==1,"MC","CR"),
    sub=ifelse(subject=="M","Math","Reading"),
    DIF=dif,
    sig=ifelse((abs(dif)/dif_se)>1.96,"sig.","non-sig.")
  ) %>%
  ggplot(aes(x=format, y=DIF)) +
  geom_hline(yintercept = 0) +
  geom_boxplot(color="grey", outlier.shape = NA, width=0.3) +
  geom_jitter(aes(color=sig), size=1, alpha=0.6, width=0.02) +
  facet_grid(sub~year) +
  xlab("") + 
  theme_bw(base_size = 14)
```

# Summary Tables

```{r}
dif %>%
  group_by(subject, year, is_mc, main_eff, theta_sd) %>%
  summarise(n=n(),
            mndif=mean(dif),
            meddif=median(dif),
            mindif=min(dif),
            maxdif=max(dif)) %>%
  mutate(is_mc=ifelse(is_mc==1,"MC","CR")) %>%
  pivot_wider(names_from = "is_mc", values_from = c("n", "mndif","meddif", "mindif","maxdif")) %>%
  mutate(format_diff=mndif_CR-mndif_MC,
         format_diff_es=format_diff/theta_sd,
         d=(main_eff*2)/theta_sd,
         n_item=n_MC+n_CR) %>%
  dplyr::select(subject, year,
                mndif_CR, meddif_CR, mindif_CR, maxdif_CR,
                mndif_MC, meddif_MC, mindif_MC, maxdif_MC,
                format_diff_es, d,
                theta_sd,
                format_diff, main_eff, n_MC, n_CR, n_item) %>%
  data.frame() %>%
  kable(digits=2) %>% kable_styling()
```

# statistical significance of DIF

```{r}
dif %>%
  mutate(
    is_sig=ifelse(abs(dif)/dif_se>1.96, 1, 0)
  ) %>%
  group_by(subject, is_mc) %>%
  summarise(
    n=n(),
    nsig=sum(is_sig),
    pos_sig=sum(dif>0 & is_sig==1),
    neg_sig=sum(dif<0 & is_sig==1)
  ) %>%
  kable() %>% kable_styling()
```

# Regression Models

## math and reading data files

```{r}
m_d <- filter(dif, subject=="M")
r_d <- filter(dif, subject=="R")
```

## Descriptive statistics

### math

```{r}
df_math <- m_d %>%
  select(dif, dif_se, is_cr, xsi_c,
         proc_emp, proc_form, proc_int_m,
         sub_chng, sub_quant, sub_space, sub_data,
         year12, year15)

sumtable(df_math)
```

### reading

```{r}
df_read <- r_d %>%
  select(dif, dif_se, is_cr, xsi_c,
         proc_acc, proc_int_r, proc_ref,
         txt_cont, txt_mix, txt_non,
         year12, year15) 

sumtable(df_read)
```

## OLS models

### Math

```{r}
m_lm1 <- lm(dif ~ is_cr + factor(year), data = m_d)
m_lm2 <- lm(dif ~ is_cr + factor(year) + xsi_c + subscale + process_short, data = m_d)
```

```{r}
tab_model(m_lm1, m_lm2, show.se=TRUE, show.ci=FALSE, string.est="B", string.se="SE")
```

#### Residuals plots

```{r}
plot(m_lm2)
```

### Reading

```{r}
r_lm1 <- lm(dif ~ is_cr + factor(year), data = r_d)
r_lm2 <- lm(dif ~ is_cr + factor(year) + xsi_c + process_short + txt_mix + txt_non, data = r_d)
```

```{r}
tab_model(r_lm1, r_lm2, show.se=TRUE, show.ci=FALSE, string.est="B", string.se="SE")
```

#### Residuals plots

```{r}
plot(r_lm2)
r_d$leverage <- hatvalues(r_lm2)
r_d %>%
  arrange(-leverage) %>%
  select(year, subject, item_id, is_mc, dif, xsi_c, process_short, text_format, leverage) %>%
  filter(leverage>0.1) %>%
  kable(digits=2) %>% kable_styling()
```

### save models

These can be imported and tabled later.

```{r}
lm_models <- list(m_lm1, m_lm2, r_lm1, r_lm2)
saveRDS(lm_models, file = "4b-lm-models-US-ypart.rds")
```

## metaregression models for Table S8

### math

```{r}
m_rma1 <- rma(yi = dif, sei = dif_se,
              mods = ~ is_cr + factor(year),
              method = "REML",
              data = m_d)
m_rma2 <- rma(yi = dif, sei = dif_se,
              mods = ~ is_cr + factor(year) + xsi_c + subscale + process_short,
              method = "REML",
              data = m_d)

summary(m_rma1)
summary(m_rma2)

tab_model(m_rma1,m_rma2,m_lm1,m_lm2,
          show.ci=FALSE,show.se=TRUE)
```

### read

```{r}
r_rma1 <- rma(yi = dif, sei = dif_se,
              method = "REML",
              mods = ~ is_cr + factor(year), data = r_d)

r_rma2 <- rma(yi = dif, sei = dif_se,
              mods = ~ is_cr + factor(year) + xsi_c + process_short + txt_non + txt_mix, 
              method = "REML",
              data = r_d)

summary(r_rma1)
summary(r_rma2)

tab_model(r_rma1,r_rma2,r_lm1,r_lm2,
          show.ci=FALSE,show.se=TRUE)
```

```{r}
tab_model(m_rma1,m_rma2,r_rma1, r_rma2,
          show.ci=FALSE,show.se=TRUE, p.style = "star",
          string.est="Est.", string.se="SE",
          dv.labels = c("Math 1", "Math 2", "Reading 1", "Reading 2"),
          pred.labels = c(
            `(Intercept)`="Intercept",
            is_cr="Is CR",
            xsi_c="Difficulty",
            subscaleQuantity="Quantity",
            `subscaleSpace and Shape`="Space and Shape",
            `subscaleUncertainty and data`="Uncertainty and Data",
            `process_shortFormulate` = "Formulate",
            `process_shortInterpret` = "Interpret",
            `process_shortIntegrate and interpret`="Integrate and Interpret",
            `process_shortReflect and evaluate`="Reflect and Evaluate",
            `txt_mix`="Text Mixed/Multiple",
            `txt_non`="Text Non-continuous"
          ),          
          file = "4b-metareg-model-table.xls")
```

